---
title: 'Python Basics (1 of 5): Getting Started with Python and Jupyter Notebooks'
output:
  html_document:
    keep_md: yes
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      eval = TRUE)
library(reticulate)
```
# Python Basics: Getting Started with Python and Jupyter Notebooks

# First time setup

I find that a lot of confusion surrounding learning a new programming language arises when you try to set up the ecosystem requiried to *work in* that language. So I will take you step by step thru the set up process assuming no familiarity with shell or command line.   

### Installing Anaconda

1. Go to [anaconda.com/download](https://www.anaconda.com/download)
2. Download the installer for your operating system (Windows, macOS, or Linux)
3. Run the installer and follow the prompts — **the default settings are fine for most users**
4. Once complete, you should be able to find **Anaconda Navigator** in your applications

> **Note for R users:** This is roughly analogous to installing R and RStudio together in a single step, with most common packages already included.

### Launching Jupyter Notebook via Anaconda Navigator

Anaconda Navigator is a point-and-click graphical dashboard — no command line required. Here is how to launch Jupyter Notebook from it:

1. Open **Anaconda Navigator** from your applications menu (search for it like any other app)
2. You will see a dashboard with several application tiles
3. Find **Jupyter Notebook** and click **Launch**
4. Jupyter will open automatically in your default web browser

> **Note:** Although Jupyter opens in your browser, it is **not running on the internet**. It is running locally on your own machine. Your files and code never leave your computer unless you explicitly share them.

This directs you towards what's called the **Jupyter file browser**, which shows your home directory. From here you can navigate to any folder and create a new **notebook**. A notebook (for those coming from a background in `R`) is analogous to an `RMD`. The file extension for a notebook is `.ipynb`. You can write text-like markdown code, write regular python code, and plot figures all in one place. The critical difference is `ipnyb` files store output in the same file, unlike `RMD` which separates the output into either HTML or PDF output in the YAML header.

### Creating Your First Notebook

1. Navigate to the folder where you want to save your work
2. Click **New** (top right) → **Python 3 (ipykernel)**
3. A new notebook will open in a new browser tab
4. Click on the title ("Untitled") at the top to rename it

Each notebook is made up of **cells**. You can run a cell by pressing `Shift + Enter`. There are three types of cells:

- **Code cells:** Where you write and execute Python code
- **Markdown cells:** Where you write formatted text, headings, and notes (change cell type via the dropdown in the toolbar)
- **Raw cells:** You can ignore these for now. They are used in advanced document conversion workflows and you're unlikely to need it in this course.

> **Note:** By default (if you're running windows) Jupyter will save the notebook to  `C:\Users\<username>\anaconda_projects`. If you're on Mac it will likely save in somewhere like `/Users/YourUsername`. You might want to change it so it saves to somewhere more familiar.

## Python vs R: Key Differences

If you are coming from `R`, you already have a strong foundation — many core concepts transfer directly. However, Python has a distinct set of conventions and behaviors that are worth understanding upfront to avoid common early frustrations.

### Philosophy and Purpose

`R` was designed from the ground up specifically for statistics and data analysis. Unlike `R`, Python is a **general-purpose language** that has become dominant in data science, but is equally at home in web development, automation, machine learning engineering, and software development more broadly.

In practice this means:

- R's syntax and built-in tools are optimized for statistical workflows
- Python's ecosystem is much broader, and its data science tools (`NumPy`, `Pandas`, `scikit-learn`) are add-ons rather than the core language itself
- Python skills transfer more easily outside of academia and into industry

### The Type System

One of the most important conceptual differences between `R` and Python is how they handle **data types**. Python is **strongly typed**, meaning it will not silently coerce one type into another. For example:

```{python, error = TRUE}
# this will throw a type error
"5" > 3
```
```{r}
# this will not
"5" > 3
```
`R`, by contrast, is much more permissive — it will often attempt to coerce types automatically, which can be convenient but sometimes leads to silent errors that are difficult to track down.

Like `R`, Python is also **dynamically typed**, meaning you do not need to declare the type of a variable upfront — it is inferred at runtime. This is the same as R. However, where R will often just "figure it out" when types are mixed, Python will stop and tell you something is wrong.

```{python}
# Python infers the type at runtime
x = 42        # int
y = 3.14      # float
z = "hello"   # str
```

### Object-Oriented vs Functional Style

R is fundamentally **functional** in style — you typically pass objects *into* functions:

```r
# R style
mean(x)
summary(df)
```

Python is fundamentally **object-oriented** — everything in Python is an object, and objects have **methods** attached to them that you call using dot `.` notation:

```{python}
import pandas as pd

data = {'x': [1, 2, 3, 4, 5]}
df = pd.DataFrame(data)

# Methods are called ON the object
df.describe()
```

Both languages can do both styles, but the default mental model differs and affects the syntax you will encounter most often.

### Whitespace is Syntax

In R, code blocks are defined with curly braces `{}` and indentation is optional (though good practice). In Python, **indentation is mandatory** — it defines the structure of your code. Getting the indentation wrong will either throw an error or, worse, silently do the wrong thing.

```{python}
# this will not error
for i in range(3):
    print(i)
```

```{python, error = TRUE}
# this will error
for i in range(3):
print(i)
```

> **Tip:** Use 4 spaces per indentation level (this is the Python convention). Most editors handle this automatically.

### Indexing: 0-based vs 1-based

This is one of the most common early stumbling blocks. In R, indexing starts at 1. In Python, it starts at 0.

```{python}
import numpy as np

my_array = np.array([10, 20, 30, 40, 50])

# First element is index 0
my_array[0]
```

```{python}
# Second element is index 1
my_array[1]
```

### Assignment Operator

- R: `x <- 10` (though `=` also works in most contexts)
- Python: `x = 10` (always)

### True/False Capitalization

- R: `TRUE` / `FALSE`
- Python: `True` / `False`

Python is case-sensitive, so `true` or `TRUE` will throw a `NameError`.

### Importing Packages

In R you install a package once and load it with `library()`. Python works similarly, but with a few differences:

- Installation: `pip install packagename` or via Anaconda (once, outside your script)
- Import: `import packagename` at the top of each script/notebook

A very common Python convention — which has no direct R equivalent — is **aliasing** packages on import. This is similar to namespace aliasing in `C++` but not quite the same:

```{python}
import numpy as np        # np is conventional
import pandas as pd       # pd is conventional
```

These aliases are not required but are so universally used in the Python data science community that you should treat them as standard.

## A bit more on Python

Python excels at interactive computational analysis, enabling extensive data exploration and manipulation. Jupyter Notebook seamlessly complements this capability, offering an interactive environment where users can write code, visualize results, and document their workflow. Through its integrated interface, Jupyter provides versatile cells for code execution, rendering outputs, and adding explanatory markdown annotations, making it a powerful tool for data analysis and scientific computing.

## Simple Math

At its most fundamental level, Python functions as a versatile computational tool, equipped to handle a wide array of mathematical operations and calculations with ease.

Addition

```{python}
2+2
```
Subtraction

```{python}
3-2
```

Multiplication

```{python}
3*3
```

Division

```{python}
4/2
```

Square root

To take the square root, we need to import the Math module first.

```{python}
import math
math.sqrt(9)
```

Log10

```{python}
math.log10(100)
```

## Functions

In Python, especially at the outset, much of your work revolves around applying functions to data. In the preceding section, we encountered various functions, including mathematical operators like `sqrt()` and `log10()`. Functions are designed to take data (or values) as input, process them, and yield an output, which is typically displayed in your console by default.

NumPy, a foundational Python library for numerical computations, introduces a crucial data type called NumPyArray. Creating these NumPy arrays becomes indispensable when collaborating with other Python libraries that heavily rely on them, such as SciPy, Pandas, Matplotlib, scikit-learn, among others. NumPy stands out for array manipulation, thanks to its wealth of built-in functions, performance optimizations, and the ability to write concise code.

To harness NumPy's capabilities, you'll first need to import the NumPy library.

```{python}
import numpy as np
```

In Python, "np" serves as a concise alias for the NumPy library. With this in place, we can proceed to generate sequential data. NumPy equips us with the "arange()" function for making arrays that encompass sequences of numbers.

```{python}
np.arange(1,11)
```

In the `arrange` function above, pay special attention to the range we entered. The end number we entered is not included in the output. This is consistent across Python so this is just something to be aware of.

## Variables

While it's convenient to view outputs in the console during interactive analysis, there are scenarios where we must retain data and outputs for future reference. This is accomplished through variable assignment, where we use the equal sign to link the values (objects) on the right with the names (variables) on the left.

```{python}
my_variable = np.arange(1,11)
```

And plug it into functions, ie, do computations on it.

```{python}
my_variable * 2
```

> When naming variables in Python, keep in mind the following:
>
> * Names should be descriptive and indicate the variable's purpose or the kind of value it holds.
> * Use lowercase letters. 
> * If a variable name consists of multiple words, separate them with underscores (e.g., `variable_name`).
> * Do not use Python keywords or functions as variable names to prevent confusion and errors.
> * Names must start with a letter or an underscore.
> * Aside from underscores, avoid using special characters in variablethe code in the long term.res.

## Data Types & Structures

Data types are fundamental constructs in Python, each with distinct characteristics. The three primary data types include numeric, character, and boolean. Numeric data allows for mathematical operations such as addition and division. Character data, often assembled into strings, consists of individual characters or groups of characters. Boolean data is crucial for handling dichotomous (true/false) values. Each data type comes with inherent properties that facilitate specific operations and manipulations.

Main Data Types:

* Numeric (int, float)
* String (str)
* Boolean (bool)

The function `type()`, will tell you what data type you have…

```{python}
type(2)
```

```{python}
type(2.2)
```

```{python}
type("a")
```

```{python}
type(True)
```

## Data Structures

Python offers various data structures to aggregate and organize multiple data elements effectively, especially in data analysis tasks. Below are some of the most frequently used data structures:

**Numpy Arrays:** These structures, provided by the NumPy library, are efficient containers that hold homogenous data (usually numbers), allowing for vectorized operations and efficient data manipulation.
  
**Pandas Series:** The Pandas library offers the Series data structure, which can hold any data type. A Series is a one-dimensional labeled array that can accommodate various data types, making it versatile for data analysis tasks.
  
**Pandas DataFrame:** Also within the Pandas library is the DataFrame, a two-dimensional, size-mutable, and potentially heterogeneous tabular data structure with labeled axes (rows and columns). It's especially suitable for handling and analyzing structured data.

Each of these structures comes with a suite of methods and attributes designed to facilitate data manipulation and analysis effectively.

We have previously provided a brief introduction to NumPy arrays. In this section, we will delve deeper into three prominent data structures, offering a more detailed exploration and understanding of each.

### NumPy Arrays

* Comes from the `numpy` library
* Homogeneous (all elements must be of the same type)
* Can be multi-dimensional (1D, 2D, 3D, ...)
* Efficient operations on numerical data.
* Indexing based on integer position.
* Supports a wide array of mathematical operations.

```{python}
import numpy as np
my_array = np.array([1, 2, 3, 4, 5])
my_array
```

```{python}
my_array*2
```

### Pandas Series

* Comes from the `pandas` library
* 1D labeled array
* Can hold any data type.
* Index can be integer-based, string-based, or even datetime-based
* Built on top of NumPy arrays 

```{python}
import pandas as pd
my_series = pd.Series([1, 2, 3, 4, 5], index=['a', 'b', 'c', 'd', 'e']) 
my_series
```

### Pandas DataFrame

* Also from the `pandas` library
* 2D labeled data structure (like a table in a database, an Excel spreadsheet, or a data frame in R)
* Can hold columns of different data types
* Built on top of NumPy arrays
* Has both row and column indices
* Most commonly used pandas object for data manipulation tasks. 

```{python}
# Define data in a dictionary
data = {
    'Name': ['John', 'Anna', 'Peter', 'Linda'],
    'Age': [28, 22, 35, 32],
    'City': ['New York', 'Paris', 'Berlin', 'Tokyo']
}

# Create DataFrame
df = pd.DataFrame(data)

# Print DataFrame
print(df)
```

* Use **NumPy arrays** when you need efficient mathematical operations on homogeneous numerical data
* Use **Pandas Series** when you need 1D labeled data
* Use **Pandas DataFrame** for most data analysis tasks, especially when working with heterogeneous tabular data.

## Getting Help within Jupyter Notebook

### Using Python's Built-in Help System

Invoke the `help()` function for information on objects, classes, functions, etc.

```{python, eval = FALSE}
help(object_name)
```

### Using Question Mark `?`

Append or prepend a `?` to an object's name for quick reference.

```{python, eval = FALSE}
object_name?
?object_name
```

For more extensive help, use double question marks `??`

```{python, eval = FALSE}
object_name??
??object_name
```

```{python, eval = FALSE}
?print
```

### Accessing Documentation Strings (docstrings)

For quick access to a function's docstring, place the cursor inside the function’s parentheses and press `Shift + Tab`.

```{python, eval = FALSE}
list()
```

### Help Menu in Toolbar

Use the "Help" menu in Jupyter Notebook's toolbar. It provides links to documentation for Jupyter, IPython, NumPy, Pandas, Matplotlib, and other libraries.

### Online Help and Searching

Search for help and documentation online by opening a new browser tab.

To get information on the Pandas `DataFrame` function, use:

```{python, eval = FALSE}
help(pd.DataFrame)
```
    


## References

**General Python Programming:**

- [Official Python Website](https://www.python.org/)
- [Python Documentation](https://docs.python.org/3/)

**Data Types in Python:**

- [Python Built-in Types](https://docs.python.org/3/library/stdtypes.html) — Official documentation on Python's built-in data types, including numeric, character (string), and boolean.

**Data Structures in Python:**

- [Python Data Structures](https://docs.python.org/3/tutorial/datastructures.html) — Overview of Python’s core data structures, with examples and usage instructions.
   
**NumPy Library and Arrays:**

- [NumPy Official Website](https://numpy.org/)
- [NumPy Documentation](https://numpy.org/doc/stable/) — Comprehensive documentation, including details on NumPy arrays and their operations.

**Pandas Library, Series, and DataFrame:**

- [Pandas Official Website](https://pandas.pydata.org/)
- [Pandas Documentation](https://pandas.pydata.org/pandas-docs/stable/) — Detailed documentation covering various functionalities, including Series and DataFrame creation and manipulation.

## Examples & Tutorials:

- [Python Basics: Data Types](https://docs.python.org/3/tutorial/introduction.html#using-python-as-a-calculator) — A basic tutorial on using various data types in Python.
- [NumPy Quickstart Tutorial](https://numpy.org/doc/stable/user/quickstart.html) — A beginner-friendly introduction to NumPy arrays and operations.
- [10 Minutes to Pandas](https://pandas.pydata.org/pandas-docs/stable/user_guide/10min.html) — A quick, 10-minute guide to getting started with Pandas, including examples of creating and working with Series and DataFrames.

> Before using any package or library, it is advisable to consult the official documentation to understand its functionalities, capabilities, and usage conventions. The official documentation is the most reliable and up-to-date source of information for Python and its libraries.
