---
title: "R: Conditions"
output:
  html_document:
    keep_md: yes
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      eval = TRUE,
                      message = FALSE,
                      warning = FALSE)
```

Last Updates: `r Sys.Date()`

Conditionally iterating over data can be a powerful tool, for such things as catching errors, testing a user input, or, perhaps more commonly in data analysis tasks, creating categorical values from numeric interval or ratio data.

## if else construct

if else statements apply a test to a value, and if it returns `TRUE`, one thing is done, and if it returns `FALSE` an alternative thing is done.

At its simplest, we can do just an if statement, which follows the syntax of

```
if (test-condition) {
  do this
  }
```

```{r}
x <- 5

if (x < 6) {
  print("x is less than 6")
}

if (x == 5) {
  print("x is equal to 5")
}

if (x != 2) {
  print("x not 2")
}
```

The if else construct allows us to string multiple test conditions together, and follows the syntax of

```
if (test-condition) {
  do this
  } else {
  do this
}
```

```{r}
if (x == 2) {
  print("x is equal to 2")
} else {
  print("x is not equal to 2")
}
```

We can begin to string things together with still more test conditions using `else if`

```
if (test-condition) {
  do this
} else if (new test-condition) {
  do this
} else if (new test-condition) {
  do this
} else {
  do this
}
```

```{r}
if (x < 5) {
  print("x is less than 5")
} else if (x > 5) {
  print("x is greater than 5")
} else {
  print("x is equal to 5")
}
```

## Iterating over vectors

A more practical implementation of an if else statement would be creating a new variable with values based off of those in another variable. This process can get a little convoluted as it combines an `if else` statement with a `for loop`.

As an example, lets assume that we want to categorize countries based on their GDP as either high, middle, or low income.

```{r}
# load the data
library(gapminder)
gapminder_cond <- gapminder

# create an empty variable to hold our ordered categorical data
gapminder_cond$income_level <- ordered(NA, levels = c("low-income",
                                                      "middle-income",
                                                      "high-income"))

# start the for loop to add values to income-level based on GDP values
for (i in seq_along(gapminder_cond$income_level)) { # for each item along the dataframe
  if (gapminder_cond$gdpPercap[i] <= 10000) { # if gdpPercap is LT = to 10000
    gapminder_cond$income_level[i] <- 'low-income' # add 'low-income' to that observation's income level variable
  } else if (gapminder_cond$gdpPercap[i] <= 75000) { # else, if gdpPercap is LT = 75000
    gapminder_cond$income_level[i] <- "middle-income" #  add 'middle-income' to that observation's income level variable
  } else { # otherwise
    gapminder_cond$income_level[i] <- "high-income"  #  add 'high-income' to that observation's income level variable
  }
}

summary(gapminder_cond$income_level)
```

This can be cumbersome. Luckily R offers some simpler options for iterating over data with an if else construct.

## ifelse()

R has a built in function, `ifelse()` that allows us to provide a condition, and, if that condition returns `TRUE` to do one thing, and if it returns `FALSE` to do something else. We'll split life expectancy into 'above average' and below average' categories.

```{r}
# Assign values
# of 'below-average' if lifeExp is equal to or below 59.47 
# and 'above-average' if above 59.47
gapminder_cond$lifeExp_cat <- ifelse(gapminder_cond$lifeExp <= 59.47, "below-average", "above-average")

# assign appropriate data type
gapminder_cond$lifeExp_cat <- ordered(gapminder_cond$lifeExp_cat, levels = c("below-average", "above-average"))

summary(gapminder_cond$lifeExp_cat)
```

`ifelse()` works well when we have a two condition set to work through. One more example, pulled from its documentation, to help avoid an error / warning being thrown or unwanted values being produced...

```{r, warning=TRUE}
# create a numeric vector
some_numbers <- c(6:-4)

# take the square root of each value, which creates a bit of an issue
sqrt(some_numbers)

# using if else to test for numbers greater than or equal to 0
# if true, calculate the square root
# if false, produce NA
sqrt(ifelse(some_numbers >= 0, some_numbers, NA))
```

## case_when()

Tidyverse's dplyr package offers an extension on `ifelse()`, allowing multiple `if else` statements.

Whereas earlier we ran

```{r, eval=FALSE}
for (i in seq_along(gapminder_cond$income_level)) { # for each item along the dataframe
  if (gapminder_cond$gdpPercap[i] <= 10000) { # if gdpPercap is LT = 10000
    gapminder_cond$income_level[i] = 'low-income' # add 'low-income' to that observation's income level variable
  } else if (gapminder_cond$gdpPercap[i] <= 75000) { # else, if gdpPercap is LT = 75000
    gapminder_cond$income_level[i] = "middle-income" #  add 'middle-income' to that observation's income level variable
  } else { # otherwise
    gapminder_cond$income_level[i] = "high-income"  #  add 'high-income' to that observation's income level variable
  }
}
```

With `case_when()`, we can simply write

```{r}
library(dplyr)

gapminder_cond$pop_size <- dplyr::case_when(
  gapminder_cond$pop <= 1000000 ~ "small",
  gapminder_cond$pop <= 100000000 ~ "medium",
  gapminder_cond$pop > 100000000 ~ "large"
)

gapminder_cond$pop_size <- ordered(gapminder_cond$pop_size, levels = c("small", "medium", "large"))

summary(gapminder_cond$pop_size)
```




