---
title: "R: Exploring Data"
output:
  html_document:
    keep_md: yes
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      eval = TRUE)
```

Last Updates: `r Sys.Date()`

## Missing Values

### is.na()

Missing values in R are coded as `NA`, a special value in R. When importing or tyding your data, it's important to ensure that you're properly identifying missing values, and handling them with the arguments `na.strings`, if using `read.csv()`, or `na`, if using `read_csv()`.

```{r}
na_values <- c("NA", "NULL", "", " ")
data_gapminder <- read.csv("../data/gapminder_nas.csv", na.strings = na_values)
```

With the data loaded, we can investigate missing values. `is.na()` returns a logical vector, `TRUE` for `NA`, otherwise `FALSE`. This is easiest with a single vector...

```{r}
# get colnames
colnames(data_gapminder)

# generate logical vector
missing_pop <- is.na(data_gapminder$pop)
typeof(missing_pop)
head(missing_pop)
```

The logical vector on it's own, is not super helpful. We can generate counts of the `TRUE` values by summing the number of `TRUE` instances in `missing_pop`...

```{r}
sum(missing_pop)
```

We can do this more interactively with

```{r}
sum(is.na(data_gapminder$pop))
```

### Complete Cases

Another point of interest in looking at missing values is to explore the number of observations for which no data are missing, which can be done with `complete.cases()`. Similar to `is.na()`, `complete.cases()` returnds a logical vector...

```{r}
complete_gapminder <- complete.cases(data_gapminder)
typeof(complete_gapminder)
```

We can get a count, like before

```{r}
sum(complete_gapminder)
```

And for use later, when subsetting, we can also generate a vector of those which are complete, identified by  their index number. This is done with `which()`

```{r}
# first few index numbers of comlpete cases
head(which(complete_gapminder))
```

## Unique values & duplicates

`unique()`, returns a vector of unique values...

```{r}
unique(data_gapminder$continent)
```

`duplicated()` returns a logical vector, `TRUE` for values (of a vector) or rows (of a data frame) that are duplicates, `FALSE` if they are unique. Since the first instance of a value is unique, the first value will return `FALSE`, and any subsequent duplicates will return `TRUE`.

```{r}
# on a vector
head(duplicated(data_gapminder$continent))

# there are 5 unique continents and 1699 duplicates
sum(duplicated(data_gapminder$continent))

# on a data frame
head(duplicated(data_gapminder))

# there are no duplicate values
sum(duplicated(data_gapminder))
```

## Descriptive Statistics

> Basic stats fucntions in R generally do not remove NA values by default, and if your data has NA values, they will return `NA`. You can avoid these errors with `na.rm = TRUE` as an argument.

### Number of Observations

The number of values in a single vector can be derived with `length()`

```{r}
length(data_gapminder$country)
```

When used on a data frame, however, `length()` returns the number of columns...

```{r}
length(data_gapminder)
```

When working with data frames, it is preferable to use `nrow()` and `ncol()` for the number of rows and columns, respectively.

```{r}
ncol(data_gapminder)
nrow(data_gapminder)
```

### Range

The range of your data can be calculated with `range()`.

```{r}
range(data_gapminder$year, na.rm = TRUE)
```

If you're only interested in the min or max values individually, this can be pulled with `min()` and `max()` respectively.

```{r}
min(data_gapminder$year, na.rm = TRUE)
max(data_gapminder$year, na.rm = TRUE)
```

### Mean & Median

The mean and median are calculated with `mean()` and `median()` respectively...

```{r}
mean(data_gapminder$lifeExp, na.rm = TRUE)
median(data_gapminder$gdpPercap, na.rm = TRUE)
```

### Variance and Standard Deviation

The variance and SD are calculated with `var()` and `sd()`...

```{r}
var(data_gapminder$lifeExp, na.rm = TRUE)
sd(data_gapminder$lifeExp, na.rm = TRUE)
```

### Quantiles

Quantiles can be derived with `quantile()`, which defaults to the max, min and interquartiles.

```{r}
quantile(data_gapminder$gdpPercap, na.rm = TRUE)
```

### Summaries

Much of the above can be gathered all at once if needed with `summary()`. `summary()` can be run on a single vector, or on an entire data frame...

```{r}
summary(data_gapminder$gdpPercap)
summary(data_gapminder)
```

A better summary would come from properly denoted data types...

```{r, echo = FALSE}
data_gapminder[1:3] <- lapply(data_gapminder[1:3], function(x) as.factor(x))
```

```{r}
# summary with the first three columns classed as categorical data
summary(data_gapminder)
```

## Visualizing Distributions

> Plotting for presentation will be covered in greater detail in a future session.

Basic plotting for distribution analysis is relatively straight forward in base R.

### Distribution of a Single Variable

#### Histograms

You can plot a basic histogram with `hist()`

```{r}
hist(data_gapminder$gdpPercap)
```

And you can increase the 'buckets' with the `breaks` argument

```{r}
hist(data_gapminder$gdpPercap, breaks = 30)
```

#### Density Plots

A density plot can be generated with `plot()`, but requires the extra step of computing the density of the data with the `density()` function first

```{r}
dens <- density(data_gapminder$lifeExp, na.rm = TRUE)
plot(dens)
```

### Distribution of Two Variables

#### Box Plots

`boxplot()` allows you to facet one variable by another using a tilde `~` to define the relationship

```{r}
boxplot(data_gapminder$lifeExp ~ data_gapminder$continent)
```

#### Scatter Plots

`plot()` by default assumes a single variable as in the earlier example, but will generate a scatter plot if provided with two variables, the first assigned to the x-axis, the second to the y-axis.

```{r}
plot(x = data_gapminder$gdpPercap, y = data_gapminder$lifeExp)
```

You can also transform you data within the function for quick investigation

```{r}
# log transform gdp per capita
plot(x = log(data_gapminder$gdpPercap), y = data_gapminder$lifeExp)
```












