---
title: 'R: Subsetting & Filtering Data'
output:
  html_document:
    keep_md: yes
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      eval = TRUE)
```

Last Updates: 2023-08-31

## Base R

### $

We've already seen that we can select a single variable of a data frame with `$`.

```{r}
data_gapminder <- read.csv("../data/gapminder.csv")
lifeExp <- data_gapminder$lifeExp
head(lifeExp)
```

It's important to note that generally R simplifies data structures when possible, so `lifeExp` has lost most of its attributes, and is now simply a vector...

```{r}
class(lifeExp)
```

`$` can be used with both data frames and lists.

```{r}
basic_list <- list(item_1 = letters,
     item_2 = 1:10,
     item_3 = c(TRUE, FALSE))

item_1 <- basic_list$item_1

item_1

# again, it defaults to creating a vector
class(item_1)
```

### [ [[

You can also subset using single or double square brackets. These provide you with more flexibility than `$`. They can also be more verbose. The primary difference between the two is that `[` preserves the original data structure, while `[[` discards it and simplifies.

```{r}
pop_preserved <- data_gapminder["pop"]
pop_simple <- data_gapminder[["pop"]]

class(pop_preserved)
class(pop_simple)
```

`[` and `[[` allow you to specify specific ranges of columns and rows to extract. The arguments are passed as `[row range, column range]`. Leaving an argument blank returns all rows or columns.

```{r}
# value in the first row and first column
data_gapminder[1,1]

# values in the second row
data_gapminder[2,]

# values in the first 5 rows and first 3 columns
data_gapminder[1:5, 1:3]
```

This works on lists as well...

```{r}
# return the first element of a list
basic_list[1]
```

However, if you want an item (value) from within a list, you need to simplify that object first with `[[` and then apply a single `[`

```{r}
# extract the first value from the first list item
basic_list[[1]][1]
```

> You can specify the row and column to subset either by its name, wrapped in "", or by its indexed number.

### Filtering Based on Conditions

You can also filter your data to specific cases based on conditions. This is done by placing the condition within the subsetting operators.

Conditions include:

| | |
| --- | --- |
| != | not equal to |
| == | equal to |
| < | less than |
| > | greater than |
| <= | less than or equal to |
| => | equal to or greater than |

And conditions can be strung together with the following operators:

| | |
| --- | --- |
| ! | not |
| \| | or |
| & | and |

#### Vectors

```{r}
# using the lifeExp vector created earlier
lifeExp_gt_70 <- lifeExp[lifeExp > 70]
head(lifeExp_gt_70, n = 20)
length(lifeExp_gt_70)
```

#### Data Frames

For data frames, you identify the column and the condition for the values in that column, which will give you all rows that match that condition, followed by the columns of interest.

```{r}
# all columns for 1952
gapminder_1952 <- data_gapminder[data_gapminder$year == 1952, ]

head(gapminder_1952)

# country and gdp for 1952

gdp_1952 <- data_gapminder[data_gapminder$year == 1952, c(1,6)]

head(gdp_1952)
```

### subset()

`subset()` allows you to achieve similar results to those above. At its most basic, it takes a data set and a condition on which to subset.

```{r}
head(subset(data_gapminder, year == 1952))
```

You can also list the columns you'd like to keep with the `select` argument.

```{r}
head(subset(data_gapminder, year == 1952,
            select = c("country", "gdpPercap")))
```

### Multiple conditions

Multiple conditions can be passed when subsetting, whether using `[` or `subset()`

```{r}
# 1952 and Americas
am_52 <- data_gapminder[data_gapminder$year == 1952 & data_gapminder$continent == "Americas", ]
head(am_52)

am_52_ss <- subset(data_gapminder, year == 1952 & continent == "Americas")
head(am_52_ss)
```

## TidyVerse

The `dplyr` package from TidyVerse draws a good conceptual break between selecting variables and filtering for cases based on values with its `select()` and `filter()` functions.

### select()

`select()` allows you pick columns based on their names. It takes two arguments, a data set, and a set of parameters by which to select columns; that parameter could be a single name, a span of names separated with a `:`, or some other condition, like names starting with, ending with, or containing specific characters, or even by the data type held in the column.

```{r, message = FALSE}
# load the library
library(dplyr)

# select 1 column
head(select(data_gapminder, country))

# select multiple columns
head(select(data_gapminder, year, lifeExp))

# select range of columns
head(select(data_gapminder, continent:pop))

# select numeric columns
head(select(data_gapminder, is.numeric))

# select by negation
head(select(data_gapminder, !country))
```

More options can be found here https://dplyr.tidyverse.org/reference/select.html 

### filter()

Filter is very similar to subset for working on data frames (it doesn't work on vectors), and follows the same basic approach of requiring a dataset and a condition on which to filter.

```{r}
head(filter(data_gapminder, continent == "Americas"))
```

`filter()` works well with other offerings from TidyVerse. And more options can be found here https://dplyr.tidyverse.org/reference/filter.html
