---
title: 'Non-parametric tests and visualizations'
output:
  html_document:
    keep_md: yes
---
# Parametric vs Non-Parametric Methods: What's the Difference?

```{r load packages, eval = TRUE, echo = FALSE}
library(knitr)
library(emoji)
```

**Parametric methods** (like t-tests and ANOVA) make strong assumptions about your data:
- Data follows a specific distribution
- Variance is equal across groups (aka homoscedasticity)
- Observations are independent
- Data are measured on an interval or ratio scale

These methods estimate **population parameters** (mean, standard deviation) and test hypotheses about those parameters.

**Non-parametric methods** (like Mann-Whitney and Kruskal-Wallis) are more flexible:
- **Distribution-free**: Don't assume a specific distribution shape
- Many work with **ranks** or frequencies instead of raw values
- More robust to outliers and skewed data
- Can handle ordinal (ranked) data
- Generally require fewer assumptions

## The Trade-off with non-parametric Tests

**Parametric tests:**

- `r emoji("heavy_check_mark")` More **powerful** (better at detecting true effects) *when assumptions are met*
- `r emoji("heavy_check_mark")` More information (estimates of means, effect sizes)
- `r emoji("skull_and_crossbones")` Can give misleading results when assumptions are violated

**Non-parametric tests:**

- `r emoji("heavy_check_mark")` **Robust** to violations of normality and outliers
- `r emoji("heavy_check_mark")` Work with small sample sizes
- `r emoji("heavy_check_mark")` Valid for ordinal data
- `r emoji("skull_and_crossbones")` Less powerful when parametric assumptions *are* met
- `r emoji("skull_and_crossbones")` Compare medians instead of means

### So When Should you Choose Non-Parametric `r emoji("thinking")`

Use non-parametric methods when:
1. **Sample size is small** (n < 30 per group) and you can't verify normality
2. **Data are heavily skewed** or have outliers
3. **Data are ordinal** (rankings, Likert scales)
4. **Normality tests fail** (Shapiro-Wilk p < 0.05)
5. **Variances are very unequal** between groups

::: {style="padding: 15px; margin: 20px 0; border-left: 5px solid #0072B2; background-color: #f0f7ff;"}
**📊 Note on Ranking and Ordinal Data**

Ordinal data are naturally ranked already ranked! Examples:
- Likert scales (e.g., 1 = Strongly Disagree, ... ,5 = Strongly Agree)
- Educational levels (High School < Bachelor's < Master's < PhD)
- Income (Low < Middle < High)

Since non-parametric tests work with ranks, they're **perfect** for ordinal data—no transformation needed!
However, much of the data that we work with day to day is not naturally ranked, so we have to
do the best we can. Thankfully, `R` provides clean ways to rank data which are not naturally **ordinal**.

```{r rank data, echo = FALSE, eval = TRUE}
n <- 10          # Number of frogs to simulate
mean_weight <- 22  # Average weight in grams
sd_weight <- 1.25  # Standard deviation in grams
frog_weights <- rnorm(n, mean = mean_weight, sd = sd_weight)
```

```{r look at ranked versus unranked data}
frog_weights_ranked <- rank(frog_weights)
comparison <- data.frame(
  Frog = 1:n,
  Weight_g = round(frog_weights, 2),
  Rank = rank(frog_weights)
)
kable(comparison, 
      col.names = c("Frog ID", "Weight (g)", "Rank"),
      caption = "Raw Weights vs Ranks",
      align = c("c", "c", "c"))
```
:::

### Rule of Thumb

If your data meet parametric assumptions, use parametric tests. If in doubt or assumptions are violated, non-parametric methods provide a safe, reliable alternative. Non-parametric methods provide powerful alternatives to parametric tests, especially useful when data do not meet the assumptions required for parametric analyses. Here’s a concise overview of the main types of non-parametric methods and their typical applications:

```{r summary table, echo = FALSE, eval = TRUE}

nonparam_summary <- data.frame(
  Test = c(
    "Mann-Whitney U\n(Wilcoxon Rank-Sum)",
    "Wilcoxon Signed-Rank",
    "Kruskal-Wallis",
    "Friedman",
    "Spearman's Correlation",
    "Chi-Square"
  ),
  
  Measures = c(
    "Difference in medians/distributions",
    "Difference in paired observations",
    "Difference in medians/distributions",
    "Difference across repeated measures",
    "Monotonic association between variables",
    "Association between categorical variables"
  ),
  
  Groups = c(
    "2 independent",
    "2 paired",
    "3+ independent",
    "3+ paired/repeated",
    "N/A (2 variables)",
    "N/A (categorical)"
  ),
  
  Uses = c(
    "Ranks",
    "Ranks (of differences)",
    "Ranks",
    "Ranks (within blocks)",
    "Ranks (of both variables)",
    "Frequency counts"
  ),
  
  Code = c(
    "wilcox.test(y ~ group)",
    "wilcox.test(x, y, paired=TRUE)",
    "kruskal.test(y ~ group)",
    "friedman.test(y ~ time | subject)",
    "cor.test(x, y, method='spearman')",
    "chisq.test(table(x, y))"
  )
)

kable(nonparam_summary, 
      caption = "Non-Parametric Tests Quick Reference",
      align = c("l", "l", "c", "c", "l"))

```

### ToothGrowth Dataset Summary

The `ToothGrowth` dataset in R examines the impact of Vitamin C on tooth growth in guinea pigs. It includes 60 observations across three variables:

- `len`: Length of odontoblasts (measured in micrometers) indicating tooth growth.
- `supp`: Type of Vitamin C supplement given (`OJ` for orange juice, `VC` for ascorbic acid).
- `dose`: Daily Vitamin C dose in milligrams (0.5, 1, and 2 mg/day).

This dataset is useful for statistical analysis, exploring how Vitamin C dosage and supplement type influence tooth growth.

```{r}
# Load the ToothGrowth data
# these data measure the effects of vitamin C on tooth growth
# in guinea pigs
data("ToothGrowth")

# Load necessary libraries
library(ggplot2)
library(dplyr)
library(tidyr)
```

```{r}
# Summary statistics
summary(ToothGrowth)
aggregate(len ~ supp + dose, data = ToothGrowth, FUN = function(x) c(mean = mean(x), sd = sd(x)))
```

```{r}
ggplot(ToothGrowth, aes(x = dose, y = len, color = supp, shape = supp)) +
  geom_point(size = 3, alpha = 0.7) +
  geom_smooth(method = "loess", se = FALSE) +
  labs(title = "Tooth Length by Dose and Supplement Type",
       x = "Dose (mg/day)",
       y = "Tooth Length",
       color = "Supplement",
       shape = "Supplement") +
  theme_minimal()
```

```{r}
# Fit the model first
model <- lm(len ~ supp, data = ToothGrowth)
# Test if RESIDUALS are normal
shapiro.test(residuals(model))
```

Shapiro-Wilk test is designed for small to medium sample sizes $50 \lt n \leq 2000$. The null hypothesis $H_0$ is that the data are normally distributed. The calculated $W$-statistic tells us the magntiude of the difference from normal, so $W == 1$ means it's normal and deviations from that (less than 1) indicate it's non-normal. 

```{r examine some residual plots}
# Test normality
# shapiro-wilk test explicitly tests for normality
qqnorm(residuals(model))
qqline(residuals(model))
```
## 1. **Comparing Two Independent Groups (Supplement Type)**
To compare the tooth length (len) between the two supplement types (supp), you can use the Mann-Whitney U test (also known as the Wilcoxon rank-sum test).

```{r}
wilcox.test(len ~ supp, data = ToothGrowth)
```
This test compares the median tooth lengths between groups given ascorbic acid (VC) and orange juice (OJ) without assuming the data are normally distributed. The null hypothesis ($H_0$) is that the median difference between the pairs is zero (no effect) and the alternative ($H_A$) is that there is a difference between the groups. `p-value` less than 0.05 indicates that they are significantly different.

## 2. **Comparing More Than Two Independent Groups (Dose Levels)**
For comparing the effects of different dose levels (0.5, 1, and 2 mg/day) on tooth length, you can use the Kruskal-Wallis test.
```{r}
kruskal.test(len ~ dose, data = ToothGrowth)
```
This test is an extension of the Mann-Whitney U test for more than two groups. It compares the median tooth lengths across different dose levels without assuming normal distribution. Here the null hypothesis reads slightly differently because we are working with 2+ groups. The null hypothesis ($H_0$) is that the median difference between all groups is zero (no effect) and the alternative ($H_A$) is that there is a difference between at least one of the groups. `p-value` less than 0.05 indicates that they are significantly different.

**While non-parametric tests do not assume normality, they still have assumptions (e.g., independence of observations). Always ensure these assumptions are met before proceeding. Additionally, non-parametric tests generally compare medians rather than means, so interpret your results accordingly.**

## Second dataset: Faithful Dataset

The `faithful` dataset comprises 272 observations of Old Faithful geyser in Yellowstone National Park, featuring:

- `eruptions`: Duration of eruptions (minutes).
- `waiting`: Waiting time until the next eruption (minutes).

```{r}
data("faithful")
summary(faithful)
hist(faithful$eruptions, breaks = 20, main = "Eruption Duration", xlab = "Duration (minutes)")
```
```{r}
model <- lm(eruptions ~ waiting, data = faithful)
# Shapiro-Wilk test for normality
shapiro.test(residuals(model))
```
Test for normality.

```{r}
qqnorm(residuals(model))
qqline(residuals(model))
```


## Suppose you want to compare the eruption durations (eruptions) between short and long waiting times. You first need to categorize the waiting times into two groups. 

```{r}
# Categorize 'waiting' into two groups
faithful$waiting_group <- ifelse(faithful$waiting < median(faithful$waiting), "Short", "Long")
# Mann-Whitney U test 
faithful_wilcox <- wilcox.test(eruptions ~ waiting_group, data = faithful)
faithful_wilcox
```

The null hypothesis ($ H_0 $) is that the median difference between the pairs is zero (no effect) and the alternative ($ H_A $) is that there is a difference between the groups. `p-value` less than 0.05 indicates that they are significantly different. Here we can clearly see that there is a significant difference ($ p $ `r faithful_wilcox$p.value`).

If you had more than two groups and wanted to compare the distributions across those groups, you’d use the Kruskal-Wallis test. For demonstration purposes, let’s simulate a scenario by dividing waiting times into three groups based on quantiles:

```{r}
# Divide 'waiting' into three groups
faithful$waiting_quantiles <- cut(faithful$waiting, breaks = quantile(faithful$waiting, probs = c(0, 1/3, 2/3, 1)), include.lowest = TRUE, labels = c("Low", "Medium", "High"))

# Kruskal-Wallis test
kruskal.test(eruptions ~ waiting_quantiles, data = faithful)

```
If you want to examine the relationship between the two variables without assuming a linear relationship or normality, use Spearman's rank correlation.

```{r}
# Spearman correlation
cor.test(faithful$eruptions, faithful$waiting, method = "spearman")
```
This means that there's strong evidence to reject the null hypothesis that there's no monotonic relationship between the two variables.

```{r}
ggplot(faithful, aes(x = waiting, y = eruptions)) +
  geom_point() +
  theme_minimal() +
  labs(title = "Correlation between Eruption Duration and Waiting Time",
       x = "Waiting Time to Next Eruption (minutes)",
       y = "Eruption Duration (minutes)")
```
```{r}
ggplot(faithful, aes(x = waiting, y = eruptions)) +
  geom_point() + 
  geom_smooth(method = "lm", color = "blue") + 
  theme_minimal() +
  labs(title = "Correlation between Eruption Duration and Waiting Time with Regression Line",
       x = "Waiting Time to Next Eruption (minutes)",
       y = "Eruption Duration (minutes)")
```

# Weaknesses of Non-Parametric Tests

Non-parametric tests are essential in statistical analysis but come with certain limitations:

- **Lower Statistical Power**: Generally, they have less power than parametric tests, making it harder to detect true effects.

- **Median Comparisons**: Often focus on medians rather than means, which may not always provide the desired insight.

- **Complex Model Limitations**: Less effective for analyzing complex relationships due to simpler modeling capabilities.

- **Reduced Parameter Information**: Provide less information about population parameters, such as mean and variance.

- **Sensitivity to Ties**: The presence of ties in data can affect the test's efficiency and discrimination ability.

- **Interpretation Challenges**: Results can be more difficult to interpret and communicate, especially to non-specialist audiences.

- **Limited Scale**: Primarily suited for ordinal or nominal data, potentially underused information from higher-level scales.